NumPy---Array Programming in Python

### 2020-02-14 body.tex
Edit discussion

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for specialized arrays to be
passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
NumPy, in turn, dispatches operations to the originating library, as required.
Over three hundred of the most popular
NumPy functions are supported.  The protocols are implemented by at
least five libraries, including Dask, CuPy, xarray, and PyData/Sparse.
Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask.
The protocols \textit{compose} well, and so it is possible to redeploy NumPy
code onto multi-GPU systems with minimal changes\cite{entschev2019}.
Working with the community, we have identified several issues that
prevent further adoption, and are refining interfaces to reduce
impedance mismatch.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
%It builds on decades of work refining the array programming
%syntax from APL through Basis, Yorick, and others.
It also makes it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
For users that require faster implementations but do not want to write lower-level
compiled code, a variety of tools exist
(e.g., Cython, Numba, Pythran) that transparently accelerate bottlenecks.
These tools allow programmers to express concepts using high-level array
programming without sacrificing efficiency.
In many cases, fast algorithms are already implemented by libraries in the
scientific Python ecosystem; these primitives simplify writing code in pure
Python.

@@

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by other
powerful interactive programming languages for scientific computing
such as Basis \cite{dubois1989basis}, Yorick \cite{munro1995using}, R \cite{ihaka1996r},
and APL \cite{iverson1962programming},
as well as commercial languages and environments like IDL and {MATLAB}.

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the broader scientific Python ecosystem and
heralded the current era of widespread use of Python for scientific computing.
%The reasons for its popularity are varied and complex but include the facts that
%NumPy (a) is built on top of Python,
%(b) has a culture of collaboration and good software development practices, and
%(c) provides a cutting-edge array programming environment for scientific data analysis.

NumPy began as an attempt to add an array object to Python and became
the foundation of a vibrant ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
rich data types such as physical units \cite{astropy,Goldbaum2018},
geometrical objects, and missing values.
% Maybe add a sentence in the direction of:
% We hope/expect, that this will allow the proliferation of rich data types
% which in turn can propel the whole ecosystem forward? (too much :)?
% As a part of other changes/developments...)
% Discussed with Jarrod, if it fits to get back to protocols, it is fine
% to call DTypes a (rich) protocol.

NumPy's role in the ecosystem continues to evolve. It is no longer just the foundational array library underlying the
scientific Python ecosystem but has become the standard API for tensor
computation and a central coordinating mechanism between array types and
technologies in Python.

>>>>

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for specialized arrays to be
passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
NumPy, in turn, dispatches operations to the originating library, as required.
Over four hundred of the most popular
NumPy functions are supported.  The protocols are implemented by at
least five libraries, including Dask, CuPy, xarray, and PyData/Sparse.
Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask.
The protocols \textit{compose} well, and so it is possible to redeploy NumPy
code onto multi-GPU systems with minimal changes\cite{entschev2019}.
Working with the community, we have identified several issues that
prevent further adoption, and are refining interfaces to reduce
impedance mismatch.

@@

NumPy combines the expressive power of \emph{array programming},
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
Many fast algorithms are already implemented by libraries in the
scientific Python ecosystem.
Where extreme optimization is warranted, compiled languages such as Cython, Numba, Pythran, that
extend Python and transparently accelerate bottlenecks, can be
used.
Because of NumPy's simple memory model, it is easy to write low-level, hand-optimized code, usually in C
or Fortran, to manipulate NumPy arrays and pass them back to
Python.
Furthermore, using array protocols, it is possible to utilize the full
spectrum of specialized hardware acceleration with minimal changes to
existing code.

@@

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that precede Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by other
powerful interactive programming languages for scientific computing
such as Basis \cite{dubois1989basis}, Yorick \cite{munro1995using}, R \cite{ihaka1996r},
and APL \cite{iverson1962programming},
as well as commercial languages and environments like IDL and {MATLAB}.

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the broader scientific Python ecosystem and
heralded the current era of widespread use of Python for scientific
computing. % This and the "What began as" sentence below are more or less interchangeable.

%What began as an attempt to add an array object to Python became
%the foundation of a vibrant ecosystem of tools.
%Now, a large amount of scientific work depends on NumPy being correct,
%fast, and stable.
%%
%%
%% v TO ADD v
%%
%% Over the next decade, the computational landscape will change
%% drastically.
%% - AI / data science mature
%% - hardware / Moore's law, specialized devices
%% - more users (diversity + numbers)
%% - scale / type of data; scientific data, lightsheet microscopes, LSST
%% - new ideas come from other languages like Julia, Rust, ...; JIT improvements
%% NumPy is poised to tackle these challenges, and work continues.
%% - Major refactoring of code
%% - Governance
%% - Process (NEP, roadmap, how we handle issues on GH, etc.)
%% - Protocols: will be extending that work
%% - Funding (some recent evidence that that is improving → grants, or
%% Moore/Sloan/CZI) CZI EOSS
%% - Redesigning website and documentation to be more friendly, easy to
%% navigate
%% - Next generation of developers (grad students) to sign on
%%

NumPy's is no longer just the foundational array library underlying
the scientific Python ecosystem, but has also become the standard API
for tensor computation and a central coordinating mechanism between
array types and technologies in Python.
